# pre-class


Make sure you commit this often with meaningful messages. 


### Getting Started


We will work with the dataset called [gapminder](https://github.com/jennybc/gapminder), this is a cleaned up version from [Gapminder Data](http://www.gapminder.org/data/). Gapminder contains a lot of great data on all of the nations of the world. We first need to install the gapminder package in R. 

```
install.packages("gapminder")
```



``` {r}
library(dplyr)
library(gapminder)
gapminder
```



### Pre-Class Problems

Use **dplyr** functions to address the following questions:

1. How many unique countries are represented per continent?

```{r}
gapminder %>% group_by(continent) %>% summarize(n_countries = n_distinct(country))
```

2. Which European nation had the lowest GDP per capita in 1997? 

```{r}
gapminder %>% filter(year == 1997 & continent == "Europe") %>% arrange(gdpPercap) # Albania
```

3. According to the data available, what was the average life expectancy across each continent in the 1980s?

```{r}
gapminder %>% group_by(continent) %>% summarise(avg_lexp = mean(lifeExp))
```

4. What 5 countries have the highest total GDP over all years combined?

```{r}
# I'm interpreting "over all years combined" to mean GDP added across years
gapminder %>% group_by(country) %>% summarise(total_gdp = sum(gdpPercap)) %>% arrange(desc(total_gdp))
# Top 5: Kuwait, Switzerland, Norway, United States, Canada

```

5. What countries and years had life expectancies of _at least_ 80 years? _N.b. only output the columns of interest: country, life expectancy and year (in that order)._

```{r}
gapminder %>% filter(lifeExp >= 80) %>% select(country, lifeExp, year)
```

6. What 10 countries have the strongest correlation (in either direction) between life expectancy and per capita GDP?

```{r}
gapminder %>% group_by(country) %>% summarise(lexp_gdp_cor = abs(cor(lifeExp, gdpPercap))) %>% arrange(desc(lexp_gdp_cor))
```

7. Which combinations of continent (besides Asia) and year have the highest average population across all countries? _N.b. your output should include all results sorted by highest average population_. With what you already know, this one may stump you. See [this Q&A](http://stackoverflow.com/q/27207963/654296) for how to `ungroup` before `arrange`ing. This also [behaves differently in more recent versions of dplyr](https://github.com/hadley/dplyr/releases/tag/v0.5.0).

```{r}
gapminder %>% filter(continent != "Asia") %>% group_by(continent, year) %>% summarise(avg_pop = mean(pop)) %>% ungroup() %>% arrange(desc(avg_pop))
# ungroup doesn't seem to be necessary here...not sure why not

```

8. Which three countries have had the most consistent population estimates (i.e. lowest standard deviation) across the years of available data? 

```{r}

gapminder %>% group_by(country) %>% summarise(sd_pop = sd(pop)) %>% arrange(sd_pop)

```


9. Subset **gm** to only include observations from 1992 and store the results as **gm1992**. What kind of object is this?

```{r}
gm1992 <- gapminder %>% filter(year == 1992)
typeof(gm1992) # list

```


10. Which observations indicate that the population of a country has *decreased* from the previous year **and** the life expectancy has *increased* from the previous year? See [the vignette on window functions](https://cran.r-project.org/web/packages/dplyr/vignettes/window-functions.html).

```{r}
gapminder %>% group_by(country) %>% summarise()
```

